---
title: "Time Series Analysis Project"
author: "Rene Lizarraga"
date: "April 21, 2019"
output:
  word_document: default
  html_document:
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Load libraries
```{r}
library(astsa)
library(fpp2)
library(ggplot2)
```


I selected "flu" dataset for analysis. Run "?flu"" in R console for data description. 


flu {astsa}	R Documentation
Monthly pneumonia and influenza deaths in the U.S., 1968 to 1978.

Description

Monthly pneumonia and influenza deaths per 10,000 people in the United States for 11 years, 1968 to 1978.

Usage

data(flu)
Format

The format is: Time-Series [1:132] from 1968 to 1979: 0.811 0.446 0.342 0.277 0.248 ...

References

http://www.stat.pitt.edu/stoffer/tsa4/


Print head of flu
```{r}
data(flu)
head(flu)
```


Verify object class is time-series. It is class ts.
```{r}
class(flu)
```


View summary statistics. 
```{r}
cat(" mean:",mean(flu))
cat(" std dev:",sd(flu))
cat(" min:",min(flu))
cat(" max:",max(flu))
```


View start, end, and frequecy of data. Data strats January 1968 and ends December 1978. The frequency is monthly. 
```{r}
cat("start:",start(flu)) 
cat(" start:",end(flu))
cat(" frequency:",frequency(flu))
```


Generate a boxplot,
There is a declining trend with an annual cycle. Flu deaths generally peak in January and bottom out in June. 
```{r}
boxplot(flu~cycle(flu))
```


Plot ab line for flu data.
```{r}
plot(flu)
abline(reg=lm(flu~time(flu)))
```


The data show a decrease of variation over time. Note the min/max spread in 1968 and 1969 versus those in subsequent years. 


Use a Box Cox lambda estimate function to estimate the degree to which the data should be transformed.
```{r}
BoxCox.lambda(flu)
```


Take this value for lambda and perform a Box Cox transformation.
```{r}
flu_trans <- BoxCox(flu,lambda = -1)
```


Plot the transformed results. This seems to reduce the variations. 
```{r}
plot(flu_trans)
```

We decompose the flu data. 

"Observed" is our transformed time series
"Trend" show the general direction up or down of a time series. In this case there is a clear downward trend.
"Seasonal" is a repeating pattern
"Random is the noise in the time series not explained by trend or seasonality. 

This plot separates the time series into the trend, seasonal, and random components. 

```{r}
plot(decompose(flu_trans))
```


Fit a regression model. The r-square of 0.12 indicates 12% of the variation is explained by the model. This result is not good. A linear regression model does not do a good job estimating the data. 

```{r}
fit_lm <- lm(flu_trans~time(flu_trans), na.action=NULL)
summary(fit_lm)
```
Plot autocorrelation (acf).  ACF exhibits a sinosoidal pattern that does not tail off and does not cut off. 
```{r}
acf(flu_trans)
```


Plot partial-autocorrelation (pacf). The PACF tails off and does not cut off lag q. 
```{r}
pacf(flu_trans)
```


Lets fit six ARIMA models


Model 1
```{r}
fit_1 <- arima(flu_trans,order = c(0,0,1))
summary(fit_1)
```


Model 2
```{r}
fit_2 <- arima(flu_trans,order = c(1,0,0))
summary(fit_2)
```


Model 3
```{r}
fit_3 <- arima(flu_trans,order = c(1,0,1))
summary(fit_3)
```


Model 4
```{r}
fit_4 <- arima(flu_trans,order = c(0,0,2))
summary(fit_4)
```


Model 5
```{r}
fit_5 <- arima(flu_trans,order = c(1,0,2))
summary(fit_5)
```


Model 6
```{r}
fit_6 <- arima(flu_trans,order = c(1,1,1))
summary(fit_6)
```


Model 5 performed best with lowest AIC of 218.01 and RMSE of 0.5285.

We will check residuals for Model 3 to see if the residuals look like white noise. 

```{r}
checkresiduals(fit_5)

```


There is still signifcant autocorrelation in the residuals and the Ljung-Box test p<0.05 show the residuals do NOT appear to be white noise. The model needs to be improved. 

I like to use the auto arima function from the R fpp2 library. 

```{r}
library(fpp2)
```


Lets fit an ARIMA model using this library. 

```{r}
fit_auto <- auto.arima(flu_trans)
summary(fit_auto)
```


The model is significantly improved with an AIC of 116.87 and RMSE of 0.3342. 

Lets check the residuals. 

```{r}
checkresiduals(fit_auto)
```


The residuals appear to pass the white noise test. The residuals do not have significant autocorrelations (except at lag-23) and the Ljung-Box test p-value is 0.118; above the alpha level of 0.05. 